Assessing primitives performance\\ on multi-stage execution

### 2018-05-17 paper.tex
Pass over all except evaluation

@@

In a hybrid runtime, where code is executed by an interpreter and a just-in-time compiler, the implementor can choose to implement optional primitives in the client language, in the virtual machine implementation language (typically C or C++), or on top of the just-in-time compiler back-end. This raises the question of the tradeoffs of the different alternatives. As a case study, we implemented the String comparison optional primitive in each case. This paper describes the different implementations and compares the execution time in Cog, a Smalltalk virtual machine.

Without surprise, the paper shows that the implementation with the most reliable and highest performance is also the most complex and difficult to implement. Nevertheless, since hundreds of optional primitives could be wished, the virtual machine implementor has to carefully choose how to implement each of them to balance between maintenance cost and high performance.
\end{abstract}

@@

For this reason, some virtual machine implementors attempted to remove most or all optional primitives. Their goal was to use a JIT performing speculative optimizations to be able to implement the optional primitives in the client language without performance loss. Ideally they wanted to get   the performance of the same primitives written in the VM. Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were the first to try this approach. Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself. However, without optional primitives implemented in the VM, several problems rise:
\begin{itemize}
	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve.
	\item The performance of the code not yet optimized is much worse than the peak performance code \sd{I did not really get it. indeed non optimized code is slower than optimized code. So what?}.
	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower. \sd{do you have a ref?}\sd{clement said: Hotspot paper}
\end{itemize}
Overall, this approach requires high engineering time to get a mature optimizing JIT with speculative optimizations and even then, it leads in practice to unreliable performance.

@@

\sd{the link is missing: In this article we want to measure the complexities and gain of this different implementations}

In this paper, we take the example of a string comparison optional primitive. The primitive compares two strings and answers if one string is greater, equal or lower than the other string. The exact specifications are detailed in Section \ref{subsec:primSpec}. We implemented a first version in Pharo/Smalltalk. We then implemented multiple version compiling through C compilers to native code, using different part of our VM infrastructure. Lastly, we implemented a version in the back-end of the JIT. Each time, we tried to write the code in the most efficient way possible. Then, we compared the execution time of all versions, showing that overall the most complex implementation is the fastest to execute.

@@

@@

\begin{figure}[bth]
		\centering
		\includegraphics[width=0.6\linewidth]{figures/VMCompilation}
		\caption{Cog VM compilation}
		\label{fig:VMCompilation}
\end{figure}

@@

Cog VM can be extended using plugins \cite{Guzd01b}.
VM plugins enable the addition of new features to the VM. There are two types of plugins:  the external and the internal ones.
\begin{itemize}
\item \emph{External plugins:} Plugins that can be added or removed from the VM without the need to recompile the VM sources.
\item \emph{Internal plugins:} Plugins that cannot be installed without recompiling the VM sources. However, these plugins can reach via an interface some parts of the VM internals such as data structures, routines,... that remains unreachable from an external plugin. This solution results in a substential overhead if called often. 
\end{itemize}

@@

The Cog VM executes the client's code using an hybrid runtime with an interpreter and a JIT. The interpreter uses a global look-up cache to speed-up virtual calls. On cache hits, it requests the JIT to compile that method and the native code version is used instead. In most case, the JIT is used at the second activation of the method. Closures have similar heuristics.

The VM code, including the interpreter code, is compiled by a C compiler and uses the C stack at runtime. During its execution, the interpreter modifies the execution stack of the client language, which is different from its own code stack as shown on Figure~\ref{fig:TwoStacks}. This means for example that the frame pointer register in the processor refers to a stack frame on the C stack and that a processor push instruction pushes a value on the C stack. When executing the native code generated by JIT, only the client language stack is used. This means for example that the frame pointer register in the processor refers to a stack frame on the client stack and that a processor push instruction pushes a value on the client stack.

@@

Because of this design, switching the execution between the interpreter and the native code generated by the JIT has a cost. Effectively, the VM needs to switch the frame pointer, the stack pointer and depending on architecture the link register \sd{I do not get "the link register"} from one stack to the other one. It also needs to save those pointers to be able to resume execution from one stack to the other one. In addition, caller-saved registers on the C conventions or the JIT convention need to be saved since the VM cannot tell ahead what code is going to be used in the other execution model and what registers are required. We estimate the overhead to up to around a dozen of processor instruction, the exact number varies a lot so it is difficult to precise it (the number of register to edit depends on the processor and on the code executed).

Most of the primitives are implemented either in Slang or directly in C. They benefit from the C compiler optimizations to be efficient. When called from the interpreter, activating a primitive is just a C call (the primitive function pointer is cached in the look-up cache next to the method to activate). When called from the native code generated by the JIT, activating a primitive requires to switch the execution from the client stack to the C stack, perform the primitive, and switch back to resume execution. If the primitive takes a significant amount of instructions to execute, for example, the primitive is copying 100kb from one array to another one, the stack switch dance overhead is negligible. If the primitive is very quick to perform, for example, the primitive is adding two integers which do not overflow, the overhead is massive: the execution cost of the addition goes from a few processor instructions to a few dozen instructions.

To solve this performance bottleneck, the Cog VM allows one to redefine primitives in Cog's Register Transfer Language (RTL), an abstract assembly which is compiled to native code through Cog's JIT back-end. Such primitives are then generated to native code at runtime when a method annotated with the primitive is compiled to native code. This allows the primitive to be performed in the client stack with the client calling convention, saving register moves and the stack switch dance. Primitives redefined in such way can be entirely or partially redefined. In the later case, only the common cases are re-implemented on top of the JIT back-end, and an unimplemented case happens at runtime, the generated native code calls the C primitive instead, adding overhead only in uncommon cases.

@@

The primitives can be associated either with a name, either with an index in the Cog VM. The named primitives can be added and modified without the need to recompile the VM.

@@

\sk{So, do we agree that: 1- we bench all implem with new spec (i am rewriting the plugins right now) 2- we drop the sections "original spec", "spec issues", rename "final spec" and rephrase the implementations details on the sections below?}
 This section first describes the different ways of implementing a primitive. We then explain how each of these versions is executed by the VM and the pros and cons of each and every.

 \subsection{Different implementations and executions}

We distinguish several implementations: the baseline in pure Smalltalk, the  \sk{delete "the first one"? sounds odd}

\paragraph{Baseline: pure Smalltalk.}
To evaluate the primitive performance, we use as baseline an implementation in pure Smalltalk. There are multiple ways of writing the comparison of two strings in Smalltalk, following Smalltalk coding conventions, a standard developer would likely use high-level constructs which apply a closure on each element of the strings. We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the constructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level. The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:. Since the focus is only on ASCII ordered comparison, we show the code only of baselineCompareWith:.

@@

\paragraph{Existing: Smart Syntax.}

The Smart Syntax Plugin is one of the version \sd{of what} existing until we started this work. The idea behind Smart Syntax is the following: the programmer writes code in Smalltalk using exclusively constructs that are recognized by the Smart Syntax compiler and that can be directly compiled to C (\emph{e.g.}, computing the size of a byte array, accessing its fields and doing integer arithmetic). With the addition of type annotations in the form of pragmas \cite{Duca16a} \footnote{Pragma are method annotation.}, the Smart Syntax plugin is able to generate C code from the Smalltalk code. \sd{A word on QuickTalk \cite{Ball86a}}
With this approach, the programmer has to write the optional primitive code a single time in Smalltalk. This code is used both as the fall-back code if the VM does not implement the primitive and at Slang to C compilation time to generate a C plugin from the code and the type annotations. This primitive follows the old specification, the full code is available in Appendix~\ref{SmartSyntaxAppendix}.

\paragraph{Existing: Plugin.}
The Slang plugin is a recent complete rewrite of the Smart Syntax plugin. The main difference is that the Smalltalk version is used only as fall-back code, while at Slang-to-C compilation time the VM uses code written in Slang to generate the C code. One of the side-effect is that the plugin's code is now part of the VM code base and not the client code bases, which is convenient for maintenance since the Cog VM has multiple client languages. However, the implementor now needs to maintain and evolve two version of the primitive, the fall-back code in Smalltalk and the code in Slang. \sd{do we need the next sentence?}This primitive follows the old specification, the full code is available in Appendix~\ref{PluginAppendix}.

\paragraph{Pure Slang.}

The Slang implementation is written inside the interpreter and has access to the full interpreter APIs. This primitive takes with two or three arguments and needs to deal with both cases in a way that the common case is efficient. The primitive is now part of the core of the VM, it cannot be easily added or removed as plugin primitives, but it is simpler to read and write. This primitive follows the new specification, the full code is available in Appendix \ref{SlangAppendix}.

\paragraph{Slang+JIT.} \sk{Id like to keep slang+JIT and not slang+RLT as JIT may ring a bell for the non-smalltalkers (while RLT may not). but as we are already using a lot of ST words (primitives, Slang...), does that actually make any sense ?}

In addition to the previous implementation, we've re-implemented here the primitive on top of Cog's JIT back-end. The primitive is written in Cog's RTL, an abstract assembly whose instructions compile almost one-to-one to native instructions. This primitive requires a more important amount of work to be written: the number of line of code is larger, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator such as Bochs. Since only performance critical implementations matter in this context, we've only re-implemented the case where there are two parameters. For the three parameter primitive, the VM falls back to the Slang version. This primitive follows the new specification, the full code is available in Appendix \ref{RTLAppendix}.

@@

\begin{table*} [bth]
\centering
\begin{tabular}{c|cccc|ccc}
   &  \multicolumn{4}{c|}{Software maintenance \& evolution} & \multicolumn{3}{c}{Primitive execution time}\\
  \hline
    & Recompilation & Familiar to & Lines & Number of & Stack switch & Optimization & Code generation \\
    & required & developers & of code & implementations & overhead & potential & control \\
  \hline
  Baseline    & No & All  & 10  & 0 & No & - & - \\
   \hline
  SmartSyntax & Plugin & VM & 20  & 3+ & Yes  & Limited & Low \\
   \hline
  SlangPlugin & Plugin & VM & 35  & 1  & Yes & Limited & Low \\
   \hline
  PureSlang   & Yes  & VM & 45  & 1  & Yes & High & Low \\
   \hline
  Slang+JIT   & Yes  & VM-JIT & 85  & 2  & No & High & High \\
  \hline
\end{tabular}
\caption{Comparison between implementations.\vspace{-0.5cm}}
\label{tab:cmp_implem}
\end{table*}

 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). \sd{why not ST, Plugin, VM?} \sk{Yes and No answers to the "recompilation needed" question, so maybe we should keep it that way}
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the approach readability, we use the number of lines of code even if this is arguable but used regularly in other papers \cite{xxx}. A better metric would be to use the cyclomatic complexity, however since the  Cog' s RTL generates branches at runtime, it is difficult to use it.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation\sd{why?}. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential on large string comparison:}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead.
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}

One or another implementation could be preferred depending on the intended aim: the main trade-off is between  engineering time and  performance. Table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time tends to increase when more low-level languages are used, such as Cog's RTL for instance.
Also, a VM recompilation affects the overall complexity of the development process, but the resulting code might be faster to run. Likewise, the use of plugins brings modularity, but as you will see in Section~\ref{sec:eval}, the performance, although better than baseline version, could still be improved.

@@

The SlangPlugin, PureSlang and Slang+JIT implementations of this string comparison primitive have been recently implemented in the Cog VM. In this section, we compare execution speed of these primitives against the other existing implementations. We measure their respective execution times for different string sizes. \sd{mention microbenchmarks here.}\sd{why there is only 4.3 and not something else? Because this is strange from a shape structure.}

@@

@@

\def\legBase{Baseline}
\def\legSmartSyntaxPlug{Syntax}
\def\legSlangPlug{Plugin}
\def\legSlang{Pure slang}
\def\legRTL{Slang+JIT}
\def\plotHeight{6.7cm}

@@

   \end{subfigure}
   \subcaption*{\textbf{LLVM-Mac}\vspace{-0.2cm}} %change vspace based on fig position
    \end{subfigure}
   \caption{String comparison benchmarks for strings of different sizes on different architectures.}
   \label{fig:bench}
\end{figure*}
\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
    & & 0 & 1 & 5 & 100 & 1000 \\
   \hline
  \multirow{2}{*}{SmartSyntax} &GCC & 0.3411 +- 0.0007 & 0.6014 +- 0.0023 & 1.3891 +- 0.0023 &  8.6796 +- 0.0587 & 13.6051 +- 0.0589\\
 							   &LLVM & 0.3147 +- 0.0023 & 0.5606 +- 0.0022 & 1.2505 +- 0.0046 & 10.3029 +- 0.0624 & 9.4065 +- 0.0367\\
   \hline
  \multirow{2}{*}{SlangPlugin} &GCC& 0.3375 +- 0.0016 & 0.5654 +- 0.0078 & 1.3150 +- 0.0205 & 8.9585 +- 0.1170  & 14.7789 +- 0.1471 \\
  							   &LLVM& 0.3036 +- 0.0018 & 0.5567 +- 0.0010 & 1.2296 +- 0.0018 & 6.6033 +- 0.0248  & 9.2138 +- 0.0140 \\
   \hline
  \multirow{2}{*}{PureSlang} &GCC& 1.0778 +- 0.0054 & 1.8169 +- 0.0191 & 3.8190 +- 0.0545 & 15.8095 +- 0.3005 & 24.4178 +- 0.1863\\
 							 &LLVM& 0.8545 +- 0.0061 & 1.5605 +- 0.0064 & 2.8205 +- 0.0080 & 10.3029 +- 0.0615  & 13.4385 +- 0.0402 \\
   \hline
  \multirow{2}{*}{Slang+JIT} &GCC& 2.4904 +- 0.0043 & 3.5378 +- 0.0124  & 6.7405 +- 0.0108 & 16.7527 +- 0.1063 & 24.4801 +- 0.0950\\
 							&LLVM & 2.2507 +- 0.0413 & 3.7545 +- 0.0111 & 7.0898 +- 0.0140 & 17.9796 +- 0.1013  & 26.4048 +- 0.0930 \\
  \hline
\end{tabular} 
\vspace{0.2cm}
\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline.}
\label{tab:detail_bench}
\end{table*}

@@

In each benchmark, we run a string comparison in a loop and measure the total execution time.
The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more). This addresses the warm-up of the JIT \cba{??? This needs to be explained differently. We get into jitted code after a small number of loop iteration (I think 20 currently) or second call. The JIT overhead is negligible here for 1 second because we could not see any time spent in the JIT in the profiler + ref to your paper, i.e. less than 0.5ms in the JIT} \cite{CogProfiler} and allows the benchmarks to run at least 1 second in the slower implementations to avoid processor / OS noises.
The loop overhead is measured and removed from the measurements.

@@

\subsection{Micro-benchmarks} \sk{Results?}

@@

We can draw the following patterns from these results:
\begin{itemize}
\item The execution time of the SmartSyntax and Plugin implementations is either very close or greater than \sd{within which ranges? 10\% 200\%} the baseline's for the small strings.
\item The Slang+JIT implementation is more efficient than the baseline in every case. Nevertheless, the performance gap between this version and the PureSlang one tends to decrease as the string size increases. 
\item The compiler used (either GCC or LLVM) does not lead to significant differences in terms of performance, except for the 100+-characters strings. 
\end{itemize}

@@

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the SmartSyntax and Plugin implementations are around 30 to 60\% slower than the baseline's, and the PureSlang performance is close to the baseline's.
As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
Regarding this criteria, the Slang+JIT version is always the fastest as it is not impacted by the switch overhead.

@@

\todo{add conclusion}

@@

%% Appendix
\null \vfill\null %switch column

\appendix
\section{Smart Syntax code}
\label{SmartSyntaxAppendix}

>>>>

\def\legBase{Baseline}
\def\legSmartSyntaxPlug{Syntax}
\def\legSlangPlug{Plugin}
\def\legSlang{Slang}
\def\legRTL{Slang+RTL}
\def\plotHeight{7cm}

@@

In a hybrid runtime, where code is executed by an interpreter and a just-in-time compiler, the implementor can choose to implement optional primitives in the client language, in the virtual machine implementation language (typically C or C++), or on top of the just-in-time compiler back-end. This raises the question of the maintenance and performance trade-offs of the different alternatives. As a case study, we implemented the String comparison optional primitive in each case. This paper describes the different implementations, discusses the maintenance cost of each of them and evaluates for different string sizes the execution time in Cog, a Smalltalk virtual machine.
\end{abstract}

@@

For this reason, some virtual machine implementors attempted to remove most or all optional primitives. Their goal was to use a JIT performing speculative optimizations to be able to implement the optional primitives in the client language without performance loss. Ideally they wanted to get   the performance of the same primitives written in the VM. Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were the first to try this approach. Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself. However, without optional primitives implemented in the VM, several problems rise:
\begin{itemize}
	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve.
	\item There is a big performance gap between the baseline performance and the peak performance.
	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower~\cite{VMHotCold}.
\end{itemize}

Overall, this approach requires high engineering time to get a mature optimizing JIT with speculative optimizations and even then, it leads in practice to unreliable performance.

@@

\paragraph{Measurements.}
In this paper, we want to measure the complexities and gain of different implementations. For this purpose, we take the example of a string comparison optional primitive. The primitive compares two strings and answers if one string is greater, equal or lower than the other string. The exact specification is detailed in Section \ref{subsec:primSpec}. The primitive is convenient for measurements since it can be executed with strings of small sizes or large sizes, leading the primitive to execute many processor instructions or only a few of them.

We implemented a first version of the optional primitive in Pharo/Smalltalk. We then implemented two versions compiling through C compilers to native code, using different parts of our VM infrastructure. Lastly, we implemented a version in the back-end of the JIT. Each time, we tried to write the code in the most efficient way possible. Then, we compared the execution time of all versions on strings of different sizes, showing that overall the most complex implementation is the fastest to execute.

@@

\begin{figure}[bth!]
		\centering
		\includegraphics[width=0.65\linewidth]{figures/VMCompilation}
		\caption{Cog VM compilation}
		\label{fig:VMCompilation}
\end{figure}

@@

@@

\paragraph{Plugins.}The Cog VM can be extended using plugins \cite{Guzd01b}. VM plugins enable the addition of new features to the VM. The main benefit of writing a plugin over extending the VM is modularity, plugins are in separate code bases and can be easily added or removed from the VM. Plugins can be compiled as internal or external:
\begin{itemize}
	\item \emph{External plugins} are compiled as a dynamic libraries distributed with the VM. They can be added or removed from a compiled VM: each dynamic library can be removed, or recompiled separately and modified.
	\item \emph{Internal plugins} are compiled with the VM executable. They can be added or removed at VM compilation time.
\end{itemize}

There is a performance overhead for the plugins compared to normal VM code. For example, accesses to object headers are memory representation dependent and therefore need to be done indirectly through calls to main VM executable, since the plugin cannot know ahead of time which memory representation is used. Such calls cannot be inlined by the C compiler\footnote{In the case of specific internal plugins, C compiler linking time optimizations can be enabled to limit this overhead.}.

@@

The Cog VM executes the client's code using an hybrid runtime with an interpreter and a JIT. The interpreter uses a global look-up cache to speed-up virtual calls. On a cache hit, it requests the JIT to compile that method and the native code version is used instead. In most case, the JIT is used at the second activation of the method. Closures have similar heuristics.

The VM code, including the interpreter code, is compiled by a C compiler and uses the C stack at runtime. During its execution, the interpreter modifies the execution stack of the client language, which is different from its own C execution stack as shown on Figure~\ref{fig:TwoStacks}. This means for example that the frame pointer register in the processor refers to a stack frame on the C stack and that a processor push instruction pushes a value on the C stack. When executing the native code generated by JIT, only the client language stack is used. This means for example that the frame pointer register in the processor refers to a stack frame on the client stack and that a processor push instruction pushes a value on the client stack.

@@

Because of this design, switching the execution between the interpreter and the native code generated by the JIT has a cost. Effectively, the VM needs to switch the frame pointer, the stack pointer and the link register (the latter only on some architectures) from one stack to the other one. It also needs to save those pointers to be able to resume execution from one stack to the other one. In addition, caller-saved registers on the C conventions or the JIT convention need to be saved since the VM cannot tell ahead what code is going to be used in the other execution model and what registers are required. We estimate the overhead to up to around a dozen of processor instructions, the exact number of numbers varies a lot (the number of registers to edit depends on the processor used and on the code executed).

Most of the primitives are implemented either in Slang or directly in C. They benefit from the C compiler optimizations to be efficient. When called from the interpreter, activating a primitive is just a C function call (the primitive function pointer is cached in the look-up cache next to the method to activate). When called from the native code generated by the JIT, activating a primitive requires to switch the execution from the client stack to the C stack, perform the primitive, and switch back to resume execution. If the primitive takes a significant amount of instructions to execute, for example, the primitive is copying 100kb from one array to another one, the stack switch dance overhead is negligible. If the primitive is very quick to perform, for example, the primitive is adding two integers which do not overflow, the overhead is massive: the execution cost of the addition goes from a few processor instructions to a few dozen instructions.

To solve this performance bottleneck, the Cog VM allows one to redefine primitives in Cog's Register Transfer Language (RTL), an abstract assembly which is compiled to native code through Cog's JIT back-end. Such primitives are then generated to native code at runtime when a method annotated with the primitive is compiled to native code. This allows the primitive to be performed in the client stack with the client calling convention, saving register moves and the stack switch dance. Primitives redefined in such way can be entirely or partially redefined. In the later case, only the common cases are re-implemented on top of the JIT back-end. If an unimplemented case happens at runtime, the generated native code calls the C primitive instead, adding overhead only in uncommon cases.

@@

%The primitives can be associated either with a name, either with an index in the Cog VM. The named primitives can be added and modified without the need to recompile the VM.

@@

This section describes the different implementations of the string comparison optional primitive and then compare them.

We then explain how each of these versions is executed by the VM and the pros and cons of each and every.

 \subsection{Different implementations}

We distinguish four implementations: the Baseline in pure Smalltalk, the Plugin version, the Slang version and the Slang+RTL version.

\paragraph{\legBase.}
To evaluate the primitive performance, we use as baseline an implementation in pure Smalltalk. There are multiple ways of writing the comparison of two strings in Smalltalk, following Smalltalk coding conventions, a standard developer would likely use high-level constructs which apply a closure on each element of the strings. We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the constructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level\footnote{The bytecode compiler moves the upper limit of the loop ahead of the loop, so (length1 min: length2) is not compute at each iteration.}. The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:. Since the focus is only on ASCII ordered comparison, we show the code only of baselineCompareWith:.

@@

%\paragraph{Existing: Smart Syntax.}
%
%The Smart Syntax Plugin is one of the version \sd{of what} existing until we started this work. The idea behind Smart Syntax is the following: the programmer writes code in Smalltalk using exclusively constructs that are recognized by the Smart Syntax compiler and that can be directly compiled to C (\emph{e.g.}, computing the size of a byte array, accessing its fields and doing integer arithmetic). With the addition of type annotations in the form of pragmas \cite{Duca16a} \footnote{Pragma are method annotation.}, the Smart Syntax plugin is able to generate C code from the Smalltalk code. 
%With this approach, the programmer has to write the optional primitive code a single time in Smalltalk. This code is used both as the fall-back code if the VM does not implement the primitive and at Slang to C compilation time to generate a C plugin from the code and the type annotations. This primitive follows the old specification, the full code is available in Appendix~\ref{SmartSyntaxAppendix}.

\paragraph{\legSlangPlug.}
The VM plugin implementation can be easily added and removed as explained in Section \ref{sec:VMCompil}. Written in Slang, it can be compiled to C independently to the VM and is then compiled to native code through the C compiler as internal or external plugin. Some accesses to the string objects (for example, checking if they are byte strings) are done indirectly through the main VM API and these calls generate some overhead. %\sd{do we need the next sentence?}This primitive follows the old specification, the full code is available in Appendix~\ref{PluginAppendix}.

\paragraph{\legSlang.}
The Slang implementation is written inside the interpreter and has access to the full interpreter APIs. This primitive takes with two or three arguments and needs to deal with both cases in a way that the common case is efficient. The primitive is now part of the core of the VM, it cannot be easily added or removed as plugin primitives, but it is simpler to read and write. %This primitive follows the new specification, the full code is available in Appendix \ref{SlangAppendix}.

\paragraph{\legRTL.} 
In addition to the previous implementation, we've re-implemented here the primitive on top of Cog's JIT back-end. The primitive is written in Cog's RTL, an abstract assembly whose instructions compile almost one-to-one to native instructions. This primitive requires a more important amount of work to be written: the number of line of code is larger, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator such as Bochs. Since only performance critical implementations matter in this context, we've only re-implemented the case where there are two parameters. For the three parameter primitive, the VM falls back to the Slang version. %This primitive follows the new specification, the full code is available in Appendix \ref{RTLAppendix}.

@@

\begin{table*} [bth]
\centering
\begin{tabular}{c|cccc|ccc}
   &  \multicolumn{4}{c|}{Software maintenance \& evolution} & \multicolumn{3}{c}{Primitive execution time}\\
  \hline
    & Recompilation & Familiar to & Lines & Number of & Stack switch & Optimization & Code generation \\
    & required & developers & of code & implementations & overhead & potential & control \\
  \hline
  \legBase    & No & All  & 10  & 0 & No & - & - \\
   \hline
  \legSlangPlug & Plugin & VM & 35  & 1  & Yes & Limited & Low \\
   \hline
  \legSlang   & Yes  & VM & 45  & 1  & Yes & High & Low \\
   \hline
  \legRTL   & Yes  & VM-JIT & 85  & 2  & No & High & High \\
  \hline
\end{tabular}
\caption{Comparison between implementations.\vspace{-0.5cm}}
\label{tab:cmp_implem}
\end{table*}

 In this section we compare the four different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). %\sd{why not ST, Plugin, VM?} %\sk{Yes and No answers to the "recompilation needed" question, so maybe we should keep it that way}
	\item \emph{Language familiar to developers:} Smalltalk developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
	\item \emph{Number of lines of code:} We measured the number of lines of code of each implementation, which arguably can be used to evaluate the complexity level of each implementation.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we need to maintain what generates the C code of the VM. We count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential on large string comparison:}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead.
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT generates the same native code whichever C compiler is used.
\end{itemize}

One or another implementation could be preferred depending on the intended aim: the main trade-off is between  engineering time and  performance. Table \ref{tab:cmp_implem} summarizes the main pros and cons related to each implementation: the engineering time tends to increase when more low-level languages are used, such as Cog's RTL for instance.
Also, a VM recompilation affects the overall complexity of the development process, but the resulting code might be faster to run. Likewise, the use of plugins brings modularity at the cost of some performance loss.

@@

\todo{Make a  pass cba}

The SlangPlugin, PureSlang and Slang+JIT implementations of this string comparison primitive have been recently implemented in the Cog VM. In this section, we compare execution speed of these primitives against the other existing implementations. We measure their respective execution times for different string sizes.% \sd{mention microbenchmarks here.}\sd{why there is only 4.3 and not something else? Because this is strange from a shape structure.}

@@

\cba{All benchs on same VM specifically compiled (normal VM has only one impl)}

@@

@@

   \end{subfigure}
   \subcaption*{\textbf{LLVM-Mac}\vspace{-0.2cm}} %change vspace based on fig position
    \end{subfigure}
   \caption{String comparison benchmarks for strings of different sizes on different architectures.}
   \label{fig:bench}
\end{figure*}
\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
    & & 0 & 1 & 5 & 100 & 1000 \\
   \hline
  \multirow{2}{*}{SlangPlugin} &GCC& 0.3375 +- 0.0016 & 0.5654 +- 0.0078 & 1.3150 +- 0.0205 & 8.9585 +- 0.1170  & 14.7789 +- 0.1471 \\
  							   &LLVM& 0.3036 +- 0.0018 & 0.5567 +- 0.0010 & 1.2296 +- 0.0018 & 6.6033 +- 0.0248  & 9.2138 +- 0.0140 \\
   \hline
  \multirow{2}{*}{PureSlang} &GCC& 1.0778 +- 0.0054 & 1.8169 +- 0.0191 & 3.8190 +- 0.0545 & 15.8095 +- 0.3005 & 24.4178 +- 0.1863\\
 							 &LLVM& 0.8545 +- 0.0061 & 1.5605 +- 0.0064 & 2.8205 +- 0.0080 & 10.3029 +- 0.0615  & 13.4385 +- 0.0402 \\
   \hline
  \multirow{2}{*}{Slang+JIT} &GCC& 2.4904 +- 0.0043 & 3.5378 +- 0.0124  & 6.7405 +- 0.0108 & 16.7527 +- 0.1063 & 24.4801 +- 0.0950\\
 							&LLVM & 2.2507 +- 0.0413 & 3.7545 +- 0.0111 & 7.0898 +- 0.0140 & 17.9796 +- 0.1013  & 26.4048 +- 0.0930 \\
  \hline
\end{tabular} 
\vspace{0.2cm}
\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline.}
\label{tab:detail_bench}
\end{table*}

@@

In each benchmark, we run a string comparison in a loop and measure the total execution time.
The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more). This addresses the warm-up of the JIT \cba{??? This needs to be explained differently. We get into jitted code after a small number of loop iteration (I think 20 currently) or second call. The JIT overhead is negligible here for 1 second because we could not see any time spent in the JIT in the profiler + ref to your paper, i.e. less than 0.5ms in the JIT} \cite{CogProfiler} and allows the benchmarks to run at least 1 second in the slower implementations to avoid processor and OS noises.
The loop overhead is measured and removed from the measurements.

@@

\subsection{Results} %\sk{Results?}

@@

We can draw the following patterns from these results:
\begin{itemize}
\item The execution time of the Plugin implementations is either very close or greater than \sd{within which ranges? 10\% 200\%} the baseline's for the small strings.
\item The Slang+JIT implementation is more efficient than the baseline in every case. Nevertheless, the performance gap between this version and the PureSlang one tends to decrease as the string size increases. 
\item The compiler used (either GCC or LLVM) does not lead to significant differences in terms of performance, except for the 100+-characters strings. 
\end{itemize}

@@

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is around 30 to 60\% slower than the baseline's, and the PureSlang performance is close to the baseline's.
As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
Regarding this criteria, the Slang+JIT version is always the fastest as it is not impacted by the switch overhead.

@@

\todo{add conclusion SK}

@@

%% Appendix
%\null \vfill\null %switch column
%
%\appendix
%\section{Smart Syntax code}
%\label{SmartSyntaxAppendix}
%
